Warming Degrees and Flowering Trees

### 2025-04-04 paper/paper.qmd
Remove unused model files and update file paths in Jupyter notebook

# Import necessary libraries 
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.dates as mdates
import seaborn as sns
import statsmodels.api as sm
import numpy as np
from matplotlib.colors import LinearSegmentedColormap
from matplotlib.ticker import MaxNLocator
from datetime import datetime
import pickle
import os

@@

@fig-modern-year-temp presents the annual full flowering dates of cherry blossoms in Japan from 1950 to 2020, measured in day-of-year (DOY), where higher values represent later blooming. Each pink dot marks the full flowering day for a given year, while the solid blue line shows a 5-year moving average, smoothing short-term fluctuations to reveal broader temporal trends. Over the 70-year span, there is a clear downward trend, indicating progressively earlier flowering.

@@

@fig-modern-temps-hist presents the frequency distribution of average March temperatures in Japan from 1950 to 2020. The temperature values are plotted on the x-axis in degrees Celsius (°C). The distribution is centered around -1.5°C, with a wider spread toward colder extremes and a thinner tail extending into positive temperatures. While most March temperatures during this period fall below freezing, the presence of warmer years, particularly in the latter decades, indicates a shift in climate norms. These warming trends align with observed shifts in cherry blossom phenology, where earlier flowering is increasingly correlated with milder spring conditions.

@@

After exploring our data, we first considered a wide variety of models to engage with before settling on using an OLS model to best interpret the data.

Initially, we considered a Cox Proportional Hazards model, which was inspired by similar applications of a Bayesian Survival model in @Allen2014. While this model seemed promising at first, there were a variety of challenges. While this model excels in handling censored data, it is less appropriate for datasets with structurally missing or entirely unrecorded entries, which characterized a large proportion of our historical data. Furthermore, the interpretive framework of the Cox Regression, which was centered around hazard ratios and survival probabilities, did not necessarily align with the explanatory goals we had in mind for this paper.

We also evaluated the possibility of using a Poisson regression model. However, the dependent variable (the day of year of sakura blooming) exhibited overdispersion during modelling, and was also normally distributed. This violates many of the core assumptions required to produce a proper interpretation from the Poisson model, rendering it statistically unsuitable for our purposes.

Ultimately, we narrowed between two candidates: the Ordinary Least Squares regression and a Regression Tree model. Given the structure of our dataset and the importance of interpretability, we selected the OLS model as our primary analytical tool. It’s coefficient-based output allowed us to quantify, interpret, and properly follow the effect of temperature and blooming times in a straightforward

@@

We are interested in investigating the relationship between sakura blooming times and the average temperature in March of the given year. The independent variable will be that of the temperature, with the dependent/responding variable being the time of blooming.

The aim of computing this OLS regression model is to estimate an equation for a line with the coefficient B1 so that the difference (error) between the predicted line and the data points are minimized. By doing this, we get an equation for a line that best fits the data, allowing for the estimation of blooming dates given the average March temperatures in a given year. With the OLS regression, we will determine and predict one single outcome variable.

@@

\begin{center}
\text{Blooming\_DoY}_i = \beta_0 + \beta_1 \cdot \text{March\_Temp}_i + \varepsilon_i
\end{center}

The dependent variable Blooming Day (DOY) represents the blooming days of Sakuras in the given year, which is measured from 0 to 365 to denote how many days of the year has passed until blooming.

The independent variable, which interacts with the slope coefficient β₁, March Temperature denotes the March Temperature of the given year, which is measured in celsius.

@@

```{python}
with open("model/ols.pkl", "rb") as ols:
    model_0 = pickle.load(ols)

with open(model/ols_robust.pkl", "rb") as ols_robust:
    model_1 = pickle.load(ols_robust)
```

@@

## Implications of Findings
Our statistical analysis reveals a significant inverse relationship between pre-flowering March temperatures and sakura bloom dates, with higher temperatures consistently associated with earlier flowering time. This temperature-dependent pattern aligns with the documented global warming trend observed over recent centuries, particularly the accelerated temperature increase in the modern era. The predictive models generated through our research demonstrate robust statistical evidence that the progressive elevation of mean March temperatures, driven by global warming, will likely continue to advance cherry blossom flowering dates in coming decades. Our findings corroborate the projections established by Allen et al. in their seminal work on phenological responses to climate change [@Allen2014]; their model, as discussed in our literature review, similarly predicted accelerated advancement of flowering dates under various warming scenarios, with comparable directional trends to those observed in our analysis. This convergence of independent methodological approaches strengthens the validity of both models and provides additional confirmation of the phenological impacts of climate alteration on various sakura species.

## Ecological Implications
The observed advancement in sakura flowering dates correlates with increasing mean March temperatures in our analysis, highlighting a significant climate-driven phenological shift. This temperature-phenology relationship aligns with a growing body of research on plant responses to climate change. As @Rahimi2025 observed in their comprehensive analysis of pollinator-dependent crops across Asia, climate change is driving significant spatial and temporal shifts in biological systems, particularly affecting species interactions that depend on synchronized timing.

The earlier flowering of sakura represents a classic example of phenological mismatch—where interacting species respond differently to climate cues, potentially disrupting ecological relationships. @Gerard2020 highlight that "many phenological shifts have already been reported at the species level for the insect emergence or blooming time of flowering plants", with temperature being the predominant driver. However, these shifts are rarely uniform across different ecological levels, creating potential desynchronization between plants and their pollinators or other dependent species.

The observed advancement in sakura blooming could have cascading effects throughout Japanese ecosystems. As @Rahimi2025 identified in their spatial mismatch analysis, when flowering phenology changes, it can disrupt the "critical synchronizations" between flowers and their respective pollinators, potentially affecting ecosystem functionality and biodiversity dynamics. In addition, not limited to spatial mismatch, Rahimi and Jung argue that temporal mismatch is also possible in the future. It will occur when the "flowering periods of plants and the activity periods of their pollinators are no longer synchronized", reducing pollination efficiency and crop yields. This phenomenon will be a likely future if global warming continues to worsen and average March temperature steadily increases.

The continued warming projected in climate models suggests that the trend toward earlier sakura flowering will likely accelerate in coming decades, potentially creating novel ecological communities as species respond differentially to changing conditions.

## Cultural Implications
From a cultural perspective, the temporal shift in cherry blossom seasons need adjustments to traditional hanami celebrations and tourism schedules, which have historically been calibrated to conventional blooming periods. In addition, these modifications could impact cultural practices that have remained relatively consistent for centuries and negatively impact tourism economy that is built around it.

## Limitations
We would like to address several methodological limiting factors. Firstly, our analysis relies predominantly on data from Kyoto, representing a geographically limited subset that may not comprehensively reflect sakura phenological patterns across Japan's diverse climate regions. Secondly, although we used a merged dataset spanning from 800 CE to 2019, there are significant data gaps/missing values where we have to drop during the cleaning & merging of the two dataset. In addition, a notable limitation is the historical dataset's reliability, as large portions comprise of reconstructed values derived from local linear regression models or originated from primary sources that lack rigorous academic verification protocols. This presents challenges for establishing data provenance and reliability in accordance with contemporary scientific standards. Lastly, while our findings highlight temperature as a statistically significant predictor of bloom timing, the single-variable approach inherently overlooks other potential ecological factors. Variables such as precipitation patterns, photoperiod fluctuations, soil conditions, and winter chilling requirements may also contribute significantly to the observed advancement in flowering dates. A more comprehensive multivariate model would potentially yield enhanced predictive capacity and ecological insights.

@@

>>>>

# Import necessary libraries 
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.dates as mdates
import seaborn as sns
import statsmodels.api as sm
import numpy as np
from matplotlib.colors import LinearSegmentedColormap
from matplotlib.ticker import MaxNLocator
from datetime import datetime
import pickle
import os
from statsmodels.iolib.summary2 import summary_col

@@

@fig-modern-year-temp above presents the annual full flowering dates of cherry blossoms in Japan from 1950 to 2020, measured in day-of-year (DOY), where higher values represent later blooming. Each pink dot marks the full flowering day for a given year, while the solid blue line shows a 5-year moving average, smoothing short-term fluctuations to reveal broader temporal trends. Over the 70-year span, there is a clear downward trend, indicating progressively earlier flowering.

@@

@fig-modern-temps-hist above presents the frequency distribution of average March temperatures in Japan from 1950 to 2020. The temperature values are plotted on the x-axis in degrees Celsius (°C). The distribution is centered around -1.5°C, with a wider spread toward colder extremes and a thinner tail extending into positive temperatures. While most March temperatures during this period fall below freezing, the presence of warmer years, particularly in the latter decades, indicates a shift in climate norms. These warming trends align with observed shifts in cherry blossom phenology, where earlier flowering is increasingly correlated with milder spring conditions.

@@

After data exploration, various modeling approaches were considered before selecting OLS regression as the most appropriate method. We were interested in investigating the relationship between sakura blooming times and the average temperature in March of the given year. The independent variable will be that of the temperature, with the dependent/responding variable being the time of blooming.

A Cox Proportional Hazards model was initially evaluated, inspired by the Bayesian Survival model used in @Allen2014. This approach was deemed unsuitable due to its emphasis on censored data rather than structurally missing entries which characterized much of the historical dataset. Additionally, the interpretive framework of hazard ratios and survival probabilities was not aligned with the explanatory objectives of this analysis.

Poisson regression was also assessed but rejected due to overdispersion in the dependent variable (day of year of sakura blooming) and its normal distribution, which violated key assumptions required for proper model interpretation.

The final consideration was between Ordinary Least Squares regression and Regression Tree models. OLS regression was ultimately selected based on the dataset structure and need for interpretability. Its coefficient-based output allowed for straightforward quantification and interpretation of the relationship between temperature and blooming times.

@@

The aim of computing this OLS regression model is to estimate an equation for a line with the coefficients β so that the differences (errors) between the predicted line and the data points are minimized. By doing this, we get an equation for a line that best fits the data, allowing for the estimation of blooming dates given the average March temperatures in a given year. With the OLS regression, we will determine and predict one single outcome variable, blooming day.

@@

$$
\text{Blooming\_DoY}_i = \beta_0 + \beta_1 \cdot \text{March\_Temp}_i + \varepsilon_i
$$

The dependent variable Blooming Day represents the blooming day of a Kyoto cherry blossoms in a given year. It is measured in days after January 1 of a given year.

The independent variable March Temperature, described by the slope coefficient $\beta_1$, denotes the mean March temperature of the given year, which is measured in celsius.

\newpage

@@

```{python}
# Load dataset
df = modern_bloom

# Rename columns for clarity
df.columns = ['year', 'flower_date', 'flower_doy', 'march_temp_c']

# Split into training (1953–2000) and testing (2010–2019)
train_df = df[df['year'] <= 2000]
test_df = df[(df['year'] >= 2009) & (df['year'] <= 2019)]

# Define X and y
X = sm.add_constant(train_df['march_temp_c'])  # Add intercept
y = train_df['flower_doy']

# Fit model 1
model_0 = sm.OLS(y, X).fit()

# Fit model 2 (with robust standard errors)
model_1 = sm.OLS(y, X).fit(cov_type='HC1')

```{python}
#| label: tbl-reg-results
#| tbl-cap: "Regression Results for Cherry Blossom Flowering Day vs. March Temperature"
#| message: false
#| echo: false
#| warning: false
#| fig-align: center

# Extract model results in a cleaner format
results_table = summary_col([model_0, model_1], 
                          model_names=["Standard OLS", "Robust OLS"],
                          float_format='%0.3f',
                          info_dict={'N':lambda x: "{0:d}".format(int(x.nobs)),
                                    'R2':lambda x: "{:.3f}".format(x.rsquared)})
# Display the table
display(results_table)
```

The regression analysis in @tbl-reg-results reveals a significant inverse relationship between March temperature and cherry blossom blooming dates. For each 1°C increase in March temperature, cherry blossoms bloom approximately 2.25 days earlier (p < 0.01). This relationship remains consistent across both standard and robust OLS models, though the latter shows larger standard errors, suggesting heteroscedasticity in the data. With an R-squared of 0.32, March temperature appears to explain about one-third of the variation in blooming dates. When March temperature is 0°C, blossoms are predicted to emerge on the 88th day of the year (late March). These findings indicate that spring temperature serves as a meaningful predictor of cherry blossom phenology.

```{python}
#| label: fig-reg-plt 
#| fig-cap: "Regression Visualization"

# Simple scatterplot with regression line
plt.figure(figsize=(8, 4))
plt.scatter(df['march_temp_c'], df['flower_doy'], color='#FF9ECD', edgecolor='#E75480', alpha=0.7)
plt.plot(sorted(df['march_temp_c']), model_0.predict(sm.add_constant(sorted(df['march_temp_c']))), color='red')
plt.text(0.1, 95, f"blooming_day = {model_0.params[1]:.2f} * temp + {model_0.params[0]:.2f}", color="red", fontsize=12)
plt.title("Sakura Blooming Day vs March Temperature")
plt.xlabel("March Avg Temperature (°C)")
plt.ylabel("Blooming Day of Year")
plt.grid(True, alpha=0.3)
```

@@

## Implications of Findings
Our statistical analysis reveals a significant inverse relationship between pre-flowering March temperatures and sakura bloom dates, with higher temperatures consistently associated with earlier flowering time. This temperature-dependent pattern aligns with the documented global warming trend observed over recent centuries, particularly the accelerated temperature increase in the modern era.

The predictive models generated through our research demonstrate robust statistical evidence that the progressive elevation of mean March temperatures, driven by global warming, will likely continue to advance cherry blossom flowering dates in coming decades. This phenological shift carries significant ecological, cultural, and economic implications. The alteration of these biorhythms may disrupt established ecological synchronizations between cherry trees and their pollinators, potentially affecting ecosystem functionality and biodiversity dynamics.

From a cultural perspective, the temporal shift in cherry blossom seasons need adjustments to traditional hanami celebrations and tourism schedules, which have historically been calibrated to conventional blooming periods. In addition, these modifications could impact cultural practices that have remained relatively consistent for centuries and negatively impact tourism economy that is built around it.

## Ecological Implications

## Cultural Implications

## Limitations

We would like to address several methodological limiting factors. Firstly, our analysis relies predominantly on data from Kyoto, representing a geographically limited subset that may not comprehensively reflect sakura phenological patterns across Japan's diverse climate regions. Secondly, although we used a merged dataset spanning from 800 CE to 2019, there are significant data gaps/missing values where we have to drop during the cleaning & merging of the two dataset. In addition, a notable limitation is the historical dataset's reliability, as large portions comprise of reconstructed values derived from local linear regression models or originated from primary sources that lack rigorous academic verification protocols. This presents challenges for establishing data provenance and reliability in accordance with contemporary scientific standards. Lastly, while our findings highlight temperature as a statistically significant predictor of bloom timing, the single-variable approach inherently overlooks other potential ecological factors. Variables such as precipitation patterns, photoperiod fluctuations, soil conditions, and winter chilling requirements may also contribute significantly to the observed advancement in flowering dates. A more comprehensive multivariate model would potentially yield enhanced predictive capacity and ecological insights.

@@

\newpage
